Evaluating FAIR Digital Object as a distributed object system

### 2023-05-18 latex/table1.tex
simplified table latex

  & The purpose of FDO is to enable federated machine actionable digital objects for scholarly purposes, in practice this also requires agreement of or compatibility between FDO types. FDO encourages research communities to develop common type registries to be shared across instances. In current DOIP practice, each service have their own types, attributes and operations. The wider symbiosis is consistent use of PIDs.
  & The Web is loosely coupled and encourages collaboration and linking by URL. In practice, REST APIs \autocite{fieldingArchitecturalStylesDesign2000a} end up being mandated centrally by dominant (often commercial) providers \autocite{fieldingReflectionsRESTArchitectural2017a}, which clients are required to use as-is with special code per service. Use of Linked Data enables common tooling and semantic mapping across differences. \\

@@

  & FDO types and operations enable detailed choreography (see CWFP). \texttt{0.TYPE/DOIPOperation} has lightweight definition of operation, \texttt{0.DOIP/Request} or \texttt{0.DOIP/Response} may give FDO Type or any other kind of ``specifics'' (incl.~human readable docs). Semantics/purpose of operations not formalised (similar operations can be grouped with \texttt{0.DOIP/OperationReference}).

>>>>

  & The purpose of FDO is to enable federated machine actionable digital objects for scholarly purposes, in practice this also requires agreement of compatibility between FDO types. FDO encourages research communities to develop common type registries to be shared across instances. In current DOIP practice, each service have their own types, attributes and operations. The wider symbiosis is consistent use of PIDs.
  & The Web is loosely coupled and encourages collaboration and linking by URL. In practice, REST APIs \autocite{fieldingArchitecturalStylesDesign2000a} end up being mandated centrally by dominant (often commercial) providers \autocite{fieldingReflectionsRESTArchitectural2017a}, and the clients are required to use each API as-is with special code per service. Use of Linked Data enables common tooling and semantic mapping across differences. \\

@@

  & FDO types and operations enable detailed choreography (Canonical Workflow Frameworks for Research \autocite{cwfr}). \texttt{0.TYPE/DOIPOperation} has lightweight definition of operation, \texttt{0.DOIP/Request} or \texttt{0.DOIP/Response} may give FDO Type or any other kind of ``specifics'' (incl.~human readable docs). Semantics/purpose of operations not formalised (similar operations can be grouped with \texttt{0.DOIP/OperationReference}).

### 2023-05-18 latex/table3.tex
simplified table latex

@@

\caption{Checking FDO guidelines \autocite{boninoFAIRDigitalObject,fdo-RequirementSpec} against its current implementations as DOIP \autocite{foundationDigitalObjectInterface} and Linked Data Platform (LDP) \autocite{FDOFramework}, with suggestions for required additions.

@@

\begin{minipage}[b]{\linewidth}\centering
\textbf{FDO Guideline}
\end{minipage}
  & \begin{minipage}[b]{\linewidth}\centering
DOIP 2.0
\end{minipage}
  & \begin{minipage}[b]{\linewidth}\centering
FDO suggestions
\end{minipage}
  & \begin{minipage}[b]{\linewidth}\centering
Linked Data Platform
\end{minipage}
  & \begin{minipage}[b]{\linewidth}\centering
LDP suggestion
\end{minipage} \\

@@

>>>>

  \begin{small}

@@

    \caption[Checking FDO guidelines against its implementations]{Checking FDO guidelines \cite{boninoFAIRDigitalObject,fdo-RequirementSpec} against its current implementations as DOIP \cite{foundationDigitalObjectInterface} and Linked Data Platform (LDP) \cite{FDOFramework}, with suggestions for required additions.

@@

  \textbf{FDO Guideline} & 
  DOIP 2.0 & 
  FDO suggestions & 
  Linked Data Platform & 
  LDP suggestion \\

@@

\end{small}

### 2023-05-18 latex/table4.tex
simplified table latex

\begin{landscape}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1642}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}

\caption[Comparing FAIR Digital Object and Web technologies as middleware infrastructures]{Comparing FAIR Digital Object with the DOIP 2.0 protocol \autocite{foundationDigitalObjectInterface} against Web technologies using Linked Data, considered as middleware infrastructures \autocite{zarrasComparisonFrameworkMiddleware2004a}

@@

Quality  & FDO with DOIP  & Web with Linked Data \\

>>>>

\begin{landscape}
\begin{small}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1642}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
	\caption[Comparing FAIR Digital Object and Web technologies as middleware infrastructures]{Comparing FAIR Digital Object (with the DOIP 2.0 protocol \cite{13TcbsZF6}) and Web technologies (using Linked Data) as middleware infrastructures \cite{hRzcHhPD}

@@

\emph{Quality} & 
FDO w/ DOIP & 
Web w/ Linked Data \\

### 2023-05-18 latex/table5.tex
simplified table latex

\caption{Assessing RDA's FAIR Data Maturity Model \autocite{groupFAIRDataMaturity2020,bahimFAIRDataMaturity2020a} (first 2 columns) against the FDO guidelines \autocite{boninoFAIRDigitalObject}, FDO implemented with the protocol DOIPv2 \autocite{foundationDigitalObjectInterface}, Linked Data Platform (LDP) \autocite{FDOFramework} and examples from Linked Data practices in general. (--- indicates \emph{Unspecified}, may be possible with additional conventions)

@@

\begin{minipage}[b]{\linewidth}\raggedright
FAIR ID
\end{minipage}
  & \begin{minipage}[b]{\linewidth}\raggedright
Indicator
\end{minipage}
  & \begin{minipage}[b]{\linewidth}\centering
FDO guidelines
\end{minipage}
  & \begin{minipage}[b]{\linewidth}\centering
FDO/DOIP
\end{minipage}
  & \begin{minipage}[b]{\linewidth}\centering
FDO/LDP
\end{minipage}
  & \begin{minipage}[b]{\linewidth}\centering
Linked Data examples
\end{minipage} \\

>>>>

    \caption[Assessing RDA's FAIR Data Maturity Model against the FDO guidelines]{Assessing RDA's FAIR Data Maturity Model \cite{1GrKTFaK2,UzQhqk0M} (first 2 columns) against the FDO guidelines \cite{RwvirqWg}, FDO implemented with the protocol DOIPv2 \cite{13TcbsZF6}, Linked Data Platform (LDP) \cite{7szz7dwO} and examples from Linked Data practices in general. (--- indicates \emph{Unspecified}, may be possible with additional conventions)

@@

  FAIR ID &
  Indicator &
  FDO guidelines &
  FDO/DOIP &
  FDO/LDP &
  Linked Data examples \\

### 2023-05-18 latex/table6.tex
converted to peerj overleaf template
Fixing references to use bibliography
more regex!
all the citations. Bioschemas by article.
simplified table latex

@@

\begin{minipage}[b]{\linewidth}\raggedright
Layer
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Recommendation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
FDO
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Linked Data
\end{minipage} \\

@@

Technical & Open Specification & FDO specifications are semi-open, process gradually more transparent & Open and transparent standard processes through W3C \& IETF \\
Technical & Common security \& privacy framework & Unspecified & TLS for encryption, multiple approaches for single-sign-on (e.g.~ORCID, Life Science Login). Privacy largely unspecified. \\
Technical & Easy SLAs for service providers & Unspecified & None \\
Technical & Access data in different formats & None formalised, custom operations or relations & Content-negotiation, \texttt{rel=alternate} relations \\
Technical & Coarse-grained/fine-grained search tools & Freetext \texttt{0.DOIP/Op.Search} on local DOIP, no federation & Coarse-grained e.g.~\href{https://datasetsearch.research.google.com/}{Google Dataset Search}, fine-grained (e.g.~federated SPARQL) require detailed vocabulary/metadata insight \\
Technical & Clear PID policy & Strong FDO requirements, tends towards Handle system. & Not required, different communities set policies \\
Semantic & Clear definitions for concepts/metadata/schemas & Required by FDO requirements, but not yet formalised & Ontologies, SKOS, OWL \\
Semantic & Semantic artefacts w/ open licenses & All artefacts are PIDs w/ license required by kernel metadata? & Open License is best practice for ontology publishing \\
Semantic & Documentation for each semantic artefact & No direct rendering from FDO, no requirement for human-readable description & Ontology rendering, content-negotiation \\
Semantic & Repositories of artefacts & Required, but not formalised & Bioontologies, etc \\
Semantic & Repositories w/ clear governance & Recommended & Largely self-governed repositories, if well-established may have clear governance. \\
Semantic & Minimal metadata model for federated discovery & Kernel metadata (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) & DCAT, ++ \\
Semantic & Crosswalks from minimal metadata model & FDO Typing recommends referencing existing type definitions, but not as separate crosswalks & Multiple crosswalks for common metadata models, but frequently not in semantic format \\
Semantic & Extensibility options for diciplinary metadata & Communities encouraged to establish own types & Extensible by design, domain-specific metadata may be at different granularity \\
Semantic & Clear protocols/building blocks for federation/harvesting of artefact catalogues & Collection types not yet defined & SWORD, OAI-PMH \\
Organisational & Interoperability-focused rules of participation recommendations & Recommended & Implied only by some communities, tendency to specialise \\
Organisational & Usage recommendations of standardised data formats & None & None -- but common for metadata (e.g.~JSON-LD) \\
Organisational & Usage recommendations of vocabularies & Recommended by community & Common (see \href{https://rdmkit.elixir-europe.org/metadata_management}{RDMKit}) \\
Organisational & Usage recommendations of metadata & Recommended by community & RO-Crate, Bioschemas \\
Organisational & Management of permanent organization names/functions & Handle owner, but unclear contact. Contact info in DOIP service provider & ROR. DCAT contacts. \\
Legal & Standardised human and machine-readable licenses & None & \href{https://spdx.org/licenses/}{SPDX} \\
Legal & Permissive licenses for metadata (CC0, CC-BY-4.0) & Undefined & Both CC0, CC-BY-4.0 common, e.g.~in DCAT \\
Legal & Different licenses for different parts & Each part as separate FDO can have separate license & DCAT, RO-Crate, Named graphs for splitting metadata \\
Legal & Mark expired/inexistent copyright & Undefined & Unclear, semantics assume copyright valid \\
Legal & Mark orphaned data & Tombstone for deleted data, but no owner of DOIP server means FDO disappears & Frequently data and endpoint has no known maintainer, archiving in common repositories becoming common \\
Legal & List recommended licenses & Undefined & Best practice recommendations \\
Legal & Track license evolution for dataset & Undefined & Versioning with PAV/PROV/DCAT \\
Legal & Policy/guidance for patent/trade secrets violation & Undefined & Undefined, legal owner may be specified \\
Legal & GDPR compliance for personal data & Undefined & Undefined \\
Legal & Restrict access/use if legally required & By transport protocol (undefined by FDO/DOIP) & Diverging approaches, typically landing pages w/ auth\&auth or click-thru \\
Legal & Harmonised terms-of-use & Undefined & Undefined \\
Legal & Alignment between EOSC and national legislation & Not applicable & Not applicable \\

>>>>

\caption[Assessing EOSC Interoperability Framework, against FDO \& Linked Data]{Assessing EOSC Interoperability Framework \autocite[section 3.6]{corchoEOSCInteroperabilityFramework2021b} against the FDO guidelines \autocite{RwvirqWg} and Linked Data practices.}
\label{tbl:eosc}\tabularnewline

@@

Layer &
Recommendation &
FDO &
Linked Data \\

@@

Technical      & Open Specification 
  & FDO specifications are semi-open, process gradually more transparent 
  & Open and transparent standard processes through W3C \& IETF \\
Technical      & Common security \& privacy framework 
  & Unspecified 
  & TLS for encryption, multiple approaches for single-sign-on (e.g.~ORCID, Life Science Login). Privacy largely unspecified. \\
Technical      & Easy SLAs for service providers 
  & Unspecified 
  & None \\
Technical      & Access data in different formats 
  & None formalised, custom operations or relations 
  & Content-negotiation, \texttt{rel=alternate} relations \\
Technical      & Coarse-grained/fine-grained search tools 
  & Freetext \texttt{0.DOIP/Op.Search} on local DOIP, no federation 
  & Coarse-grained e.g.~\href{https://datasetsearch.research.google.com/}{Google Dataset Search}, fine-grained (e.g.~federated SPARQL) require detailed vocabulary/metadata insight \\
Technical      & Clear PID policy 
  & Strong FDO requirements, tends towards Handle system. 
  & Not required, different communities set policies \\
Semantic       & Clear definitions for concepts/metadata/schemas 
  & Required by FDO requirements, but not yet formalised 
  & Ontologies, SKOS, OWL \\
Semantic       & Semantic artefacts w/ open licenses 
  & All artefacts are PIDs, license not yet required by kernel metadata
  & Open License is best practice for ontology publishing \\
Semantic       & Documentation for each semantic artefact 
  & No direct rendering from FDO, no requirement for human-readable description 
  & Ontology rendering, content-negotiation \\
Semantic       & Repositories of artefacts 
  & Required, but not formalised 
  & Bioontologies, otherwise not usually federated \\
Semantic       & Repositories w/ clear governance 
  & Recommended 
  & Largely self-governed repositories, if well-established may have clear governance. \\
Semantic       & Minimal metadata model for federated discovery 
  & Kernel metadata \autocite{fdo-KernelAttributes} based on RDA recommendations \autocite{weigelRDARecommendationPID2018}.
  & DCAT, schema.org, Dublin Core \\
Semantic       & Crosswalks from minimal metadata model 
  & FDO Typing recommends referencing existing type definitions, but not as separate crosswalks 
  & Multiple crosswalks for common metadata models, but frequently not in semantic format \\
Semantic       & Extensibility options for diciplinary metadata 
  & Communities encouraged to establish own types 
  & Extensible by design, domain-specific metadata may be at different granularity \\
Semantic       & Clear protocols/building blocks for federation/harvesting of artefact catalogues 
  & Collection types not yet defined 
  & SWORD, OAI-PMH \\
Organisational  & Interoperability-focused rules of participation recommendations 
  & Recommended 
  & Implied only by some communities, tendency to specialise \\
Organisational  & Usage recommendations of standardised data formats 
  & None 
  & None -- but common for metadata (e.g.~JSON-LD) \\
Organisational  & Usage recommendations of vocabularies 
  & Recommended by community 
  & Common (see \href{https://rdmkit.elixir-europe.org/metadata_management}{RDMKit}) \\
Organisational  & Usage recommendations of metadata 
  & Recommended by community 
  & RO-Crate, Bioschemas \\
Organisational  & Management of permanent organization names/functions 
  & Handle owner, but unclear contact. Contact info in DOIP service provider 
  & ROR. DCAT contacts. \\
Legal          & Standardised human and machine-readable licenses 
  & None 
  & \href{https://spdx.org/licenses/}{SPDX}, but not that frequently used \\
Legal          & Permissive licenses for metadata (CC0, CC-BY-4.0) 
  & Undefined 
  & Both CC0, CC-BY-4.0 common, e.g.~in DCAT \\
Legal          & Different licenses for different parts 
  & Each part as separate FDO can have separate license 
  & DCAT, RO-Crate, Named graphs for splitting metadata \\
Legal          & Mark expired/inexistent copyright 
  & Undefined 
  & Unclear, semantics assume copyright valid \\
Legal          & Mark orphaned data 
  & Tombstone for deleted data, but no owner of DOIP server means FDO disappears 
  & Frequently data and endpoint has no known maintainer, archiving in common repositories becoming common \\
Legal          & List recommended licenses 
  & Undefined 
  & Best practice recommendations \\
Legal          & Track license evolution for dataset 
  & Undefined 
  & Versioning with PAV/PROV/DCAT \\
Legal          & Policy/guidance for patent/trade secrets violation 
  & Undefined 
  & Undefined, legal owner may be specified \\
Legal          & GDPR compliance for personal data 
  & Undefined 
  & Undefined \\
Legal          & Restrict access/use if legally required 
  & By transport protocol (undefined by FDO/DOIP) 
  & Diverging approaches, typically landing pages w/ auth\&auth or click-thru \\
Legal          & Harmonised terms-of-use 
  & Undefined 
  & Undefined \\
Legal          & Alignment between EOSC and national legislation 
  & Not applicable 
  & Not applicable \\
